Alternative Methods for Implementing Explicit and Finding Implicit Sharing in embedded DSLs

### 2022-12-15 paper/paper.tex
finished core body, needs polishing, abstract, intro, conclusion

\section{Implicit Sharing Via ByteString ASTs}
TODO

\subsection{Memory Limitations}
TODO
\section{Explicit Sharing Of ByteString ASTs}
TODO

@@

% LocalWords:  DSLs unsafePerformIO Haskell's AST typeclass tagless
% Local Variables:
% LaTeX-verbatim-environments-local: ("minted")
% eval: (setq-local LaTeX-indent-environment-list (cons '("minted" current-indentation) (default-value 'LaTeX-indent-environment-list)))
% End:

>>>>

\section{Implicit Sharing Via ByteString ASTs}

The heart of our problem is whenever we need to sequence the state of the inputs
for one of our DSL functions we want to first check if it's already been
evaluated. But how do we do that without first evaluating it to gain access to
it's unique NodeID. We need some other way to uniquely identify it.

Our proposed solution is too build an AST using byte strings along with our
DAG, but hold it outside of the State monad. We can then build our DAG using a
Trie, using the byte string AST to lookup our node identifiers instead.
% TODO reference literature on tries

\begin{minted}{haskell}
data Graph a = Graph { unGraph :: State DAG NodeID
                     , stringAST :: ByteString }

data DAG = DAG { Trie (Node,NodeID)
               } deriving Show
\end{minted}

The Trie essentially serves as our new BiMap, the resulting DAG contained in its
lookup values. We essentially still perform the hash-consing technique but using
the AST to perform the lookup
\begin{minted}{haskell}
hashcons :: ByteString -> Node -> State DAG NodeID
hashcons sAST node = do
 DAG trie maxID <- get
 case Trie.lookup sAST trie of
   Nothing -> let maxID' = maxID+1
                  trie' = Trie.insert sAST (node,maxID+1) trie
               in do put $ DAG trie' maxID'
                     return maxID'
   Just (_,nodeID) -> return nodeID

instance Exp Graph where
  constant x = let
    node = NConstant x
    sAST = buildStringAST node []
    in Graph (hashcons sAST $ NConstant x) sAST
  variable x = let
    node = NVariable x
    sAST = buildStringAST node []
    in Graph (hashcons sAST $ NVariable x) sAST
  add e1 e2 = let
      sAST = buildStringAST (NAdd undefined undefined) [unStringAST e1,unStringAST e2]
      sT = do ns <- seqArgs [e1,e2]
              case ns of
                [n1,n2] -> hashcons sAST $ NAdd n1 n2
                _ -> error "black magic"
    in Graph sT sAST

\end{minted}

The instance implementations for \mintinline{haskell}{constant} and
\mintinline{haskell}{variable} work roughly the same, the novelty of the method
is in how we handle DSL functions that take other DSL State as input like
\mintinline{haskell}{add}. First we need to construct a byte string AST from
it's input ASTs, there's a lot of ways we could go about this to attempt to
minimize memory. A naive implementation would look similar to a pretty printer.
Then we need to sequence it's inputs without evaluating the inner state if
unnecessary. We do this through the implementation of \mintinline{haskell}{seqArgs}
\begin{minted}{haskell}
seqArgs :: [Graph a] -> State DAG [NodeID]
seqArgs inps =
  let
    seqArg (Graph sT sAST) =
      do DAG trie _ <- get
         case Trie.lookup sAST  trie of
           Nothing -> sT
           Just (_,nodeID) -> return nodeID
  in sequence $ map seqArg inps
\end{minted}

We only evaluate the inner state \mintinline{haskell}{sT} of each argument if we
fail to look up its corresponding byte string AST in the Trie. This will prevent
redundant hashconsing without the need for explicit sharing. However this method
suffers from its own drawbacks.

\subsection{Memory Limitations}
The byte string AST being built will itself suffer from lack of sharing. We're
essentially trading extra computation for extra memory. This is often a good
tradeoff, since memory is so plentiful in modern hardware. But under the right
conditions it can become an issue

TODO include heap profiling analysis

\section{Explicit Sharing Of ByteString ASTs}
We propose another solution to this issue, taking inspiration again from the
\cite{kiselyov:sharing}, we can introduce an explicit construct for specifying
sharing. This time, the construct will substitute the current byte string for a
more compact label. For safety purposes, we need to keep track of a table of
these labels and their corresponding ASTs, to make sure we don't insert of the
same labels.
\begin{minted}{haskell}
data DAG = DAG { dagTrie :: Trie (Node,NodeID)
               , dagCacheMap :: Map ByteString ByteString
               } deriving Show

data Graph a = Graph { unGraph :: State DAG NodeID
                     , stringAST :: ByteString
                     , addCache :: Maybe ByteString }

class Cacheable repr where
  cache :: ByteString -> repr ByteString -> repr ByteString
instance Cacheable Graph where
  cache s' (Graph g s _) = Graph g s' (Just s)

-- TODO better example?
test x y = let
  z = cache "z" (add x y)
  in add z z
\end{minted}

The cache operation replaces the current byte string AST with a new label, and
we'll define a new operation \mintinline{haskell}{runCache} that will check if
the label already exists in the cache map before inserting it.
\begin{minted}{haskell}
runCache :: ByteString -> Maybe ByteString -> Trie ByteString -> State DAG ()
runCache sAST mAddCache cacheMap = do
  case mAddCache of
    Nothing -> return ()
    Just sAST0 ->
      case Trie.lookup sAST cacheMap of
        Nothing -> let cacheMap' = Trie.insert sAST sAST0 cacheMap
                   in modify (\dag -> dag { dagCache = cacheMap' })
        Just sAST1 -> if sAST1 == sAST0
                         then return ()
                         else error $ "attempted to recache: " ++ show sAST

seqArgs :: [Graph a] -> State DAG [NodeID]
seqArgs inps =
  let
    seqArg (Graph sT sAST mCache) =
      do DAG trie cacheMap _ <- get
         runCache sAST mCache cacheMap
         ...
\end{minted}
We need to make sure we don't attempt to insert the same cache label for two
different ASTs. Unfortunately, if there is a collision there's no way to escape
the State monad to prevent or modify the subtitution. The best we can do is
crash the program, or if we use a monad tansformer, we ould use
Control.Monad.Except to through an exception. Either way it's up to the DSL
writer to insure they don't reuse the same label.

\section{Conclusion}
TODO

@@

% LocalWords:  DSLs unsafePerformIO Haskell's AST typeclass tagless Trie
% Local Variables:
% LaTeX-verbatim-environments-local: ("minted")
% eval: (setq-local LaTeX-indent-environment-list (cons '("minted" current-indentation) (default-value 'LaTeX-indent-environment-list)))
% End:
